The yaq Project: \\ Standardized Software Enabling Flexible Instrumentation

### 2023-03-16 SI.tex
non-blocking

Script writers may simply wait for ``is\_busy'' to return ``false''.

\begin{codefragment}{Python}
import yaqc
import time
motor = yaqc.Client(port=38000)
motor.set_position(500)  # initiate a long motor motion
while motor.is_busy():
    time.sleep(0.001)
print("done")  # will only print when motion is complete
\end{codefragment}

@@

\begin{codefragment}{Python}
import time
import yaqc
motor1 = yaqc.Client(port=38000)
motor2 = yaqc.Client(port=38001)
sensor = yaqc.Client(port=38002)
data = []
for m1 in range(-10, 10, 1):
    for m2 in range(0, 300, 5):
        motor1.set_position(m1)
        motor2.set_position(m2)
        for c in [motor1, motor2]:
            while c.busy():
                time.sleep(0.001)
        sensor.measure()
        while sensor.busy():
            time.sleep(0.001)
        reading = dict()
        reading["timestamp"] = time.time()
        reading["motor1"] = motor1.get_position()
        reading["motor2"] = motor2.get_position()
        reading.update(sensor.get_measured())
        data.append(reading)
\end{codefragment}

@@

@@

The plot tab (Figure \ref{acq:fig:plot_tab}) presents a live representation of recent data collected for the current plan.
The primary window is a PyQtGraph\cite{pyqtgraph} plot showing the most recent five slices of collected data.
The current slice is the brightest cyan slice, with the previous slices shown as duller colors progressively until the last one which is gray.
At the top, the currently plotted channel name and most recent collected value are shown in large font size.
The plotting does handle array detectors, for which the top number shown is the maximum value of the most recent array collected.
At present, cameras and other higher dimensional sensors are not able to be plotted using the live plot tab.
On the right hand side, there are controls to adjust the plot.
At the top, a selector allows for choosing which channel to plot.
The second selector determines which values appear along the X-axis of the graph.
Additionally, there is a selector which allows for changing the units of the X-axis.
Finally, there is a Scan Index indicator, which indicates which indexed pixel was the most recent collected.

@@

The last major change to \yaq{}'s protocol format was the adoption of Apache Avro-RPC\cite{AvroSpecification}.
Avro is similarly terse like Msgpack for serializing messages, while having a well defined specification for defining an RPC schema: a static definition of available messages, including their signatures and documentation.
This statically defined schema pairs well with \yaq{}'s trait system.
Daemon authors provide a short, minimal TOML file which describes the implemented traits and unique messages for a particular daemon.
This then gets transformed by an automated tool to a full Avro-RPC JSON file describing the protocol.
Avro is then used to encode individual messages with minimal overhead.

@@

yaqd-control can be installed via
PyPI\cite{yaqd-control} or
conda-forge\cite{yaqd-control-conda}.

@@

The raw data for package size analysis was collected using Tokei\cite{tokei}, a command line tool for analysing line lengths of source code.
It includes breaking down line length into ``code'', ``comments'', and ``blank'' lines.
The package size data was curated into \texttt{figures/lines\_histogram.txt}, which includes annotations as to the ``type'' and ``class'' for each file.
The ``type'' annotation is one of ``stub'' (meaning that the bulk of the implementation is in another file), ``normal'' (an individual daemon with the bulk of its implementation), ``protocol'' (an implementation the provides for multiple daemons, such as those referred to by ``stub''s, or ``serial'' (the implementation of a serial protocol rather than a full fledged daemon).
The ``class'' refers to the primary function of the daemon and is one of the following values: ``is-sensor'' for detectors, ``has-position'' for setable hardware, ``serial'' (identical to ``type'' annotation), and ``other'' for daemons which do not fit in the above categories.

@@

>>>>

Script writers may simply wait for ``busy'' to return ``false''.

\begin{codefragment}{Python}
import yaqc
import time
motor = yaqc.Client(port=38000)
motor.set_position(500)  # initiate a long motor motion
while motor.busy():
    time.sleep(0.001)
print("done")  # will only print when motion is complete
\end{codefragment}

@@

\begin{codefragment}{Python}
import time
import yaqc
motor1 = yaqc.Client(port=38000)
motor2 = yaqc.Client(port=38001)
sensor = yaqc.Client(port=38002)
data = []
for m1 in range(-10, 10, 1):
    motor1.set_position(m1)
    for m2 in range(0, 300, 5):
        motor2.set_position(m2)
        for c in [motor1, motor2]:
            while c.busy():
                time.sleep(0.001)
        sensor.measure()
        while sensor.busy():
            time.sleep(0.001)
        reading = dict()
        reading["timestamp"] = time.time()
        reading["motor1"] = motor1.get_position()
        reading["motor2"] = motor2.get_position()
        reading.update(sensor.get_measured())
        data.append(reading)
\end{codefragment}

@@

\subsection{Non-Blocking Script}

\begin{codefragment}{Python}
import time
import yaqc
temp = yaqc.Client(port=38003)
sensor = yaqc.Client(port=38004)
data = []
# begin heating
temp.set_position(100)
# take data until temp greater than 90 deg
while temp.get_position <= 90:
    sensor.measure()
    while sensor.busy():
        time.sleep(0.001)
    reading = dict()
    reading["timestamp"] = time.time()
    reading["temperature"] = temp.get_position()
    reading.update(sensor.get_measured())
    data.append(reading)
\end{codefragment}

@@

The plot tab (Figure \ref{acq:fig:plot_tab}) presents a live representation of recent data collected for the current plan.
The primary window is a PyQtGraph \cite{pyqtgraph} plot showing the most recent five slices of collected data.
The current slice is the brightest cyan slice, with the previous slices shown as duller colors progressively until the last one which is gray.
At the top, the currently plotted channel name and most recent collected value are shown in large font size.
The plotting does handle array detectors, for which the top number shown is the maximum value of the most recent array collected.
At present, cameras and other higher dimensional sensors are not able to be plotted using the live plot tab.
On the right hand side, there are controls to adjust the plot.
At the top, a selector allows for choosing which channel to plot.
The second selector determines which values appear along the X-axis of the graph.
Additionally, there is a selector which allows for changing the units of the X-axis.
Finally, there is a Scan Index indicator, which indicates which indexed pixel was the most recent collected.

@@

The last major change to \yaq{}'s protocol format was the adoption of Apache Avro-RPC \cite{AvroSpecification}.
Avro is similarly terse like Msgpack for serializing messages, while having a well defined specification for defining an RPC schema: a static definition of available messages, including their signatures and documentation.
This statically defined schema pairs well with \yaq{}'s trait system.
Daemon authors provide a short, minimal TOML file which describes the implemented traits and unique messages for a particular daemon.
This then gets transformed by an automated tool to a full Avro-RPC JSON file describing the protocol.
Avro is then used to encode individual messages with minimal overhead.

@@

yaqd-control can be installed via
PyPI \cite{yaqd-control} or
conda-forge \cite{yaqd-control-conda}.

@@

The raw data for package size analysis was collected using Tokei \cite{tokei}, a command line tool for analysing line lengths of source code.
It includes breaking down line length into ``code'', ``comments'', and ``blank'' lines.
The package size data was curated into \texttt{figures/lines\_histogram.txt}, which includes annotations as to the ``type'' and ``class'' for each file.
The ``type'' annotation is one of ``stub'' (meaning that the bulk of the implementation is in another file), ``normal'' (an individual daemon with the bulk of its implementation), ``protocol'' (an implementation the provides for multiple daemons, such as those referred to by ``stub''s, or ``serial'' (the implementation of a serial protocol rather than a full fledged daemon).
The ``class'' refers to the primary function of the daemon and is one of the following values: ``is-sensor'' for detectors, ``has-position'' for setable hardware, ``serial'' (identical to ``type'' annotation), and ``other'' for daemons which do not fit in the above categories.

@@

\section{The name \yaq{}}

The name \yaq{} started out as an ``inside joke'' with several meanings. \\
For posterity, we describe those meanings here.
\begin{enumerate}
  \item \yaq{} is similar to the popular acronym ``daq'' (Data AcQuisition)
  \item \yaq{} could stand for Yet Another acQuisition, a reference to the popular tongue-in-cheek naming convention ``yet another X'' in software development \cite{Raymond_1996}. This is a particularly poignant reference for the authors, who are well aware that \yaq{} is ``yet another project'' in a long line of increasingly sophisticated hardware interface software packages in the Wright Group.
  \item \yaq{} can reference the animal ``yak'', \textit{Bos grunniens}, who serves as a mascot for the project.
  \item \yaq{} is an obscure reference to the term ``yak shaving'', coined by Carlin Vieri at MIT in the 1990s \cite{CarlinVieri}. Quoting from Wiktionary \cite{yak_shaving}: yak shaving is \textit{any apparently useless activity which, by allowing you to overcome intermediate difficulties, allows you to solve a larger problem}. The entire \yaq{} project is one giant ``yak shaving'' task that ultimately enables progress towards our primary mission---creative, reproducible, insightful experimental science.
\end{enumerate}

### 2023-03-16 paper.tex
citation cleanup (#57)
non-blocking

\title{The yaq Project: \\ Standardized Software Enabling Flexible Instrumentation}
\author{Kyle F. Sunden}
\author{Daniel D. Kohler}
\author{Kent A. Meyer}
\author{Peter L. Cruz Parrilla}
\author{John C. Wright}
\author{Blaise J. Thompson}
\affiliation{
  University of Wisconsin--Madison
}
\email{blaise.thompson@wisc.edu}

@@

As smaller research labs have grown in experimental complexity, many individual labs have created domain-specific orchestration software.
In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity\cite{WeberSebastien2021a,Bogdanowicz2022,trspectrometer,Koerner_2019,Campagnola_2014,pymeasure,Giesbrecht_2022,pylablib}.
While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
Most small custom research instrumentation continues to rely on monolithic software which has hard-coded interface support for each particular connected device.
These monolithic applications tend to be inflexible and difficult to develop.

We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
We have built this standard to be self-describing, portable, and reusable wherever possible.
Via the \yaq{} interface, we provide easy-to-use hardware support for scientists to use with their choice of experiment orchestration software.

@@

At the bottom, we graphically represent the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
Bluesky has no built-in hardware interface support, instead relying on packages to create hardware-interfaces that are compatible with their Hardware Protocol defintions.
\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} is a specialized client which adapts any \yaq{} protocol into Bluesky's Hardware Protocol.
Similar translation layers could be built for a variety of orchestration software such as PyMoDAQ\cite{WeberSebastien2021a}, PyMeasure\cite{pymeasure}, or TRSpectrometer\cite{trspectrometer}.

@@

Full set of line-count data and script used to produce Figure \ref{fig:histogram}.

@@

The authors would like to thank all \yaq{} users and contributors.
We would also like to acknowledge the developers of the broader open source software community upon-which this project rests.

@@

\section*{References}
\nocite{*}
\bibliography{references}

>>>>

\title{The yaq Project: \\ Standardized Software Enabling Flexible Instrumentation}
\author{Kyle F. Sunden}
\author{Daniel D. Kohler}
\author{Kent A. Meyer}
\author{Peter L. Cruz Parrilla}
\author{John C. Wright}
\author{Blaise J. Thompson*}
\affiliation{
  University of Wisconsin--Madison;
  Madison, Wisconsin, 53706;
  USA
}
\email{blaise.thompson@wisc.edu}

@@

As smaller research labs have grown in experimental complexity, many individual labs have created domain-specific orchestration software.
In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity\cite{WeberSebastien2021a,Bogdanowicz2022,trspectrometer,Koerner_2019,Campagnola_2014,pymeasure,Giesbrecht_2022,pylablib}.
While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
More work needs to be done to catalog, compare, and contrast the large number of open-source projects that now exist.
Most small custom research instrumentation continues to rely on monolithic software which has hard-coded interface support for each particular connected device.
These monolithic applications tend to be inflexible and difficult to develop.

We have created a new network-based communication standard for scientific instrumentation, \yaq{} (Yet Another acQuisition)\footnote{see Supplementary Material for a full description of the name ``\yaq{}''}.
This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
We have built this standard to be self-describing, portable, and reusable wherever possible.
Via the \yaq{} interface, we provide easy-to-use hardware support for scientists to use with their choice of experiment orchestration software.

@@

At the bottom, we graphically represent the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
Bluesky has no built-in hardware interface support, instead relying on packages to create hardware-interfaces that are compatible with their Hardware Protocol defintions.
\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} is a specialized client which adapts any \yaq{} protocol into Bluesky.
Similar translation layers could be built for a variety of orchestration software such as PyMoDAQ\cite{WeberSebastien2021a}, PyMeasure\cite{pymeasure}, or TRSpectrometer\cite{trspectrometer}.

@@

Full set of line-count data and script used to produce Figure \ref{fig:histogram}.
More detailed examples of orchestration using \yaq{}, including scripts and graphical user interfaces.
Detailed description of \texttt{yaqd-control}.
Discussion and quantitative analysis of the \yaq{} interface's performance with large arrays.

@@

The authors would like to thank all \yaq{} users and contributors.
We would also like to acknowledge the developers of the broader open-source software community upon-which this project rests.

@@

\section*{References}
\bibliography{references}
